Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-24 optim.qmd
edit section 4 and 5

The data structure proposed in @RJ-2021-105 is implemented for JS
optimisers and additional information such as index function used, data
dimension, and jellyfish hyper-parameters can be included as additional
columns. An example data collected from finding the sine-wave structure
in 6D data (`d = 6`) using a spline index function(`splines2d`) is
printed below:

@@

The jellyfish hyper-parameters used are 50 jellyfishes (`n_jellies`) and
a maximum of 50 tries (`max_tries`). The full data contains 50
simulations, recorded by the variable `sim`, with different seeds
(`seed`). Recorded in columns `basis`, `index_val`, `tries`, `loop` and
`time`, are the sampled projection basis, calculated index value,
iteration ID, jellyfish ID, and time taken to find the optimum,
respectively. The basis column records every visited basis before
comparing with the current (last) basis. If a visited basis has a
smaller index value than the current (last) one, JSO will keep the
current basis for the next iteration in the search but still records the
visited inferior basis. This data can be used to derive numerical
summaries and plots:

-   **Success rate**: The success rate is defined as the proportion of
    simulations that achieve a final index value within 0.05 of the best
    index value found among all 50 simulations. In the dataset above,
    the optimal index value is
    `r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`.
    Out of the 50 simulations,
    `r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb`
    achieved an index value within \[`r aa - 0.05`, `r aa`\], resulting
    in a success rate of
    `r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`.
    @fig-projection shows the best projection found in each simulation,
    sorted by the index value. Most of the projections show a clear
    sine-wave structure, while the last six show only a straight line
    with clusters at the end. When local optima exist in the projection
    pursuit problem, plotting the projections found in each simulation
    or by each jellyfish can help identify these local optima.

-   **Projection distance**: The projection distance between projection
    bases and the theoretical best basis (or the empirical best basis if
    the theoretical best is unknown) is calculated as the Frobenius
    norm, which measures how close a visited basis is to the best basis.
    Plotting this projection distance against the index value for each
    basis visited reveals how the index value changes as the basis
    approaches the optimal basis, helping to identify the squintability
    of the index function. @fig-index-value-proj-dist shows the index
    value against the projection distance for the given example and its
    binned version. As the projection distance decreases, the index
    value initially increases with an increasing slope and then levels
    off. The index value shows greater variation at medium projection
    distances compared to small and large distances.

```{r}
#| label: fig-projection
#| fig.height: 6
#| fig.width: 10
#| fig.cap: Best projections found by the JS opitmiser in each simulation, sorted by the index value. The projection pursuit problem is to detect the 6D sine-wave structure using the spline index. Most simulations have detected a clear sine-wave pattern, while the last six show a straight line with clustering at the ends.. In this simulation, the success rate is 0.88 (44/50).
proj_dt <- map_dfr(sim_sine_6d_spline_best |> pull(basis), 
              ~as_tibble(as.matrix(sine1000) %*% .x), .id = "sim") |> 
  mutate(sim = as.numeric(sim)) |> 
  left_join(sim_sine_6d_spline_best |> select(sim, index_val), by = "sim")

@@

The JSO provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search-better optimiser and explores how it behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in @sec-theory,
along with jellyfish hyper-parameters, on the outcome of the
optimisation.

@@

Fifty simulations are conducted with both JSO and the search-better , in
four dimensions ($d = 6, 8, 10, 12$). The JSO uses 100 jellyfishes and
100 maximum number of tries, while the search-better optimiser allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. @fig-proj presents the final projections found by the two
optimisers, broken down by 10th quantile, faceted by the data dimension.
In the 6-dimensional data scenario, JSO consistently identifies a clear
pipe shape. The search-better optimiser also finds the pipe shape but
with a wide rim, suggesting a further polish search may be required.
With increasing dimensions, JSO may not always identify the pipe shape
due to random sampling, but it still presents a pipe shape in over 50%
of cases. When compared to the search-better optimiser, JSO reaches
higher index values and clearer pipe shapes across all quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of JSO . The hyper-parameters tested include: 1) 20, 50, and 100
jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in @sec-vis.
@fig-proportion presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the optimal projection.

@@

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, along with the jellyfish search hyper-parameters (the
number of jellyfishes and the maximum number of tries), affect the
success of JSO. Simulations are conducted to obtain the success rate
across various projection pursuit problems, characterised by
shape-to-find, data dimension, and index function used, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
@sec-theory. A generalised linear model is used to construct the
relationship between the success rate, jellyfish hyper-parameters and
optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
`dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining
with two jellyfish hyper-parameters, a total of `r nrow(sim_df)` cases
is produced, comprising of `r nrow(filter(sim_df, index == "holes"))`
pipe-finding cases and `r nrow(filter(sim_df, index != "holes"))`
sine-wave finding cases. For each case, JSO is run fifty times and the
success rate is calculated as in @sec-app-1.

Smoothness and squintability are computed for each case, following the
procedures outlined in @sec-smoothness and @sec-squintability. To
calculate smoothness, three hundred random bases are simulated. Index
values and projection distance (to the optimal basis) are calculated for
each random basis before fitting them into a Gaussian process model. For
squintability, fifty random bases are sampled and interpolated to the
optimal basis with a step size of 0.005. For these interpolated bases,
index values and projection distances to the optimal basis are
calculated and averaged with a bin width of 0.005. The squintability
measure is then calculated from fitting a 4-parameter scaled logistic
function of index value against projection distance, estimated by
non-linear least squares.

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as equation
(@eq-squintability-parametric) as squintability. The low
squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

@@

@tbl-mod-data combines all the variables calculated above (the JSO
success rate, JSO hyper-parameters, and optimisation properties) into
one table. Three preprocessing steps are applied to the data: 1) scaling
the JSO hyper-parameters by a factor of 10 for interpretation, 2)
creating a new binary variable `long_time` to indicate cases with an
average run time over 30 seconds, and 3) encoding the success rate for
the stringy case as 0, because upon visual inspection of the final
projection, none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and @tbl-mod-output presents the model
outputs. The model suggests that the JSO success rate is positively
associated with the two JSO hyper-parameters, as well as with the
optimisation properties -- smoothness and squintability. Specifically,
using 10 more jellyfish and 10 more tries will increase the odd ratio of
success by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`%
and `r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension will reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. Users can increase the number of jellyfish and the number of
tries to improve the success rate, until JSO exceeds a 30-second
runtime.

>>>>

The data structure proposed in @RJ-2021-105 is implemented for JSO. 
An example data collected from finding the sine-wave structure
in 6D data using a spline index function (`splines2d`) is
printed below:

@@

The JSO hyper-parameters used are 50 jellyfishes (`n_jellies`) and
a maximum of 50 tries (`max_tries`). The full data contains 50
simulations, indexed by the variable `sim`, with the associated `seed`s. 
Recorded in columns `basis`, `index_val`, `tries`, `loop` and
`time`, are the sampled projection basis, index value,
iteration ID, jellyfish ID, and computational time
respectively. The basis column records every visited basis before
comparing with the current (last) basis. If a visited basis has a
smaller index value than the current (last) one, JSO will keep the
current basis for the next iteration in the search but still records the
visited inferior basis. This data can be used to derive numerical
summaries and plots:

-   **Success rate**: The success rate is defined as the proportion of
    simulations that achieves a final index value within 0.05 of the best
    index value found among all 50 simulations. In the dataset above,
    the optimal index value is
    `r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`.
    Out of the 50 simulations,
    `r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb`
    achieved an index value within \[`r aa - 0.05`, `r aa`\], resulting
    in a success rate of
    `r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`.
    @fig-projection shows the best projection found in each simulation,
    sorted by the index value. Most of the projections show a clear
    sine-wave structure, while the last six show only a straight line
    with clusters at the end. When local optima exist in the projection
    pursuit problem, plotting the projections found in each simulation
    or by each jellyfish can help identify these local optima.

-   **Projection distance**: The projection distance between projection
    bases and the optimal best basis (or the empirical best basis if
    the optimal is unknown) measures how close a visited basis is to 
    the best basis and it is calculated as the Frobenius norm in Equation
    @eq-proj-dist.
    Plotting projection distances against index values for the 
    bases visited reveals how the index value changes as the basis
    approaches the optimal basis, helping to calculate the squintability
    measure. @fig-index-value-proj-dist shows the index
    value against the projection distance for the given dataset and its
    binned version. As the projection distance decreases, the index
    value initially increases with an increasing slope and then levels
    off. The index value shows greater variation at medium projection
    distances compared to small and large distances.

```{r}
#| label: fig-projection
#| fig.height: 6
#| fig.width: 10
#| fig.cap: Best projections found by the JSO in each of the 50 simulations, sorted by index values. The projection pursuit problem is to detect the 6D sine-wave structure using the spline index. Most simulations have detected a clear sine-wave pattern, while the last six show a straight line with clustering at the ends. In this simulation, the success rate is 0.88 (44/50).
proj_dt <- map_dfr(sim_sine_6d_spline_best |> pull(basis), 
              ~as_tibble(as.matrix(sine1000) %*% .x), .id = "sim") |> 
  mutate(sim = as.numeric(sim)) |> 
  left_join(sim_sine_6d_spline_best |> select(sim, index_val), by = "sim")

@@

The JSO provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares the 
performance of JSO with an exist optimiser, search better, and explores how JSO behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in @sec-theory,
along with JSO hyper-parameters, on the outcome (success rate) of the
optimisation.

@@

Fifty simulations are conducted with both JSO and the search better optimal, in
four dimensions ($d = 6, 8, 10, 12$). JSO uses 100 jellyfishes and
100 maximum number of tries, while the search-better optimiser allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. @fig-proj presents the final projections found by the two
optimisers, broken down by 10th quantile, faceted by the data dimension.
In the 6-dimensional data scenario, JSO consistently identifies a clear
pipe shape. The search better optimiser also finds the pipe shape but
with a wide rim, suggesting a further polish search may be required.
With increasing dimensions, JSO may not always identify the pipe shape
due to random sampling, but it still presents a pipe shape in over 50%
of cases. When compared to the search better optimiser, JSO reaches
higher index values and clearer pipe shapes across all quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the JSO success
rate. The hyper-parameters tested include: 1) 20, 50, and 100
jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in @sec-vis.
@fig-proportion presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the projection close to the optimal.

@@

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of
projection pursuit optimisation problems. This example investigates how
these metrics, along with the hyper-parameters affect the JSO
success rate. Simulations are conducted to obtain the success rate
across a wider array of projection pursuit problems, characterised by
shape-to-find, data dimension, and the index function, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as described in
@sec-theory. A generalised linear model is used to construct the
relationship between the success rate, JSO hyper-parameters and
optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indexes considered:
`dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining
with two JSO hyper-parameters, a total of `r nrow(sim_df)` cases
is produced, comprising of `r nrow(filter(sim_df, index == "holes"))`
pipe-finding cases and `r nrow(filter(sim_df, index != "holes"))`
sine-wave finding cases. For each case, JSO is run fifty times and the
success rate is calculated as in @sec-app-1.

Smoothness and squintability are computed for each case, following the
procedures outlined in @sec-smoothness and @sec-squintability. To
calculate smoothness, three hundred random bases are simulated. Index
values and projection distance (to the optimal basis) are calculated for
each random basis before fitting into a Gaussian process model to obtain 
the smoothness measure. For
squintability, fifty random bases are sampled and interpolated to the
optimal basis with a step size of 0.005. Index values and projection 
distances are calculated for these interpolated bases and the index values 
are averaged with a bin width of 0.005. The squintability
measure is then calculated from fitting a 4-parameter scaled logistic
function of index value against projection distance, estimated by
non-linear least squares.

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as equation @eq-squintability-parametric as squintability. The low
squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

@@

@tbl-mod-data combines all the variables calculated above (JSO
success rate, hyper-parameters, and optimisation properties) into
one table. Three pre-processing steps are applied to the data: 1) scaling
the JSO hyper-parameters by a factor of 10 for interpretation, 2)
creating a new binary variable `long_time` to indicate cases with an
average run time over 30 seconds, and 3) encoding the success rate for
the stringy case as 0, because upon visual inspection of the final
projection, none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and @tbl-mod-output presents the model
outputs. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the
optimisation properties -- smoothness and squintability. Specifically,
using 10 more jellyfishes and 10 more tries will increase the odd ratio of
success by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`%
and `r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension will reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. Users should consider increasing the number of jellyfish and the number of
tries to improve the success rate, until JSO exceeds a 30-second
runtime.
